A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-11 podc_paper/algorithm-GC.tex
final edits

\gc{
\Function{Block}{SplitBlock}{\typ{Node} v}
	\State \linecomment{figure out where to split \var{v}'s RBT}
	\If{$\var{v}=\var{root}$}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, \var{v}.\fld{last}[k])$
		\EndFor
		\State $B\assign \var{root}.\fld{blocks}[m-1]$
	\Else
		\State $B_p \assign \Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
		\State \var{dir} \assign ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B \assign \var{v}.\fld{blocks}[B_p.\edir]$
	\EndIf
	\State \linecomment If $B$ was discarded, use leftmost block instead
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \label{substitute}
\EndFunction{SplitBlock}
}

@@

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
    \If {$\var{v} = \var{root}$} \Return{true}
    \Else
    	\State $T\assign \var{v}.\fld{parent.blocks}$
		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
		\If{\Call{MaxBlock}{$T$}.\edir < b} \Return \fa
		\Else
			\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper3}
			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
		\EndIf
	\EndIf
\EndFunction{Propagated}
}

@@

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{351}
\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
    	\State \linecomment{ queue is empty when dequeue occurs}
        \gc{
        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
        \EndIf
        }
        \State \Return \nl \hfill \label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
		\gc{
        \State \var{res} \assign \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b_e$
        \EndIf
        \State \Return \var{res}
		}
    \EndIf
\EndFunction{FindResponse}

@@

>>>>

\gc{
\Function{Block}{SplitBlock}{\typ{Node} v}
	\State \linecomment{figure out where to split \var{v}'s RBT}
	\If{$\var{v}=\var{root}$}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, \var{v}.\fld{last}[k])$
		\EndFor
		\State $B\assign \var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
	\Else
		\State $B_p \assign \Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
		\State \var{dir} \assign ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B \assign \var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
	\EndIf
	\State \linecomment If $B$ was discarded, use leftmost block instead
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \label{substitute}
\EndFunction{SplitBlock}
}

@@

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
    \If {$\var{v} = \var{root}$} \Return{true}
    \Else
    	\State $T\assign \var{v}.\fld{parent.blocks}$
		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa
		\Else
			\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper3}
			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
		\EndIf
	\EndIf
\EndFunction{Propagated}
}

@@

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{351}
\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRnumenqGC}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
    	\State \linecomment{ queue is empty when dequeue occurs}
        \gc{
        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
        \EndIf
        }
        \State \Return \nl \hfill \label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
		\gc{
        \State \var{res} \assign \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b_e$
        \EndIf
        \State \Return \var{res}
		}
    \EndIf
\EndFunction{FindResponse}

@@

In the code, we use $\var{v}.\fld{blocks}[i]$ to refer to the block in the RBT stored in $\var{v}.\fld{blocks}$
with index~$i$.
A search for this block may sometimes not find it, if it has already been discarded
by another process's GC.
As mentioned in Section \ref{reducing}, if this happens to an enqueue operation,
the enqueue can simply terminate because the fact that the block is gone means that another
process has helped the enqueue reach the root of the ordering tree.
Similarly, if  a dequeue operation performs a failed search on a RBT, the dequeue can return the value 
written in the \fld{response} field of the 
leaf block that represents the dequeue and terminate, since some other process
will have written the \fld{response} there before discarding the needed block.
We do not explicitly write this early termination every time we do a lookup in an RBT.
There is one exception to this rule:  if an RBT lookup for block $B$ returns \nl\
on line \ref{SBlookup1} or \ref{SBlookup2} of
\op{SplitBlock} because the required block has been discarded, 
we continue doing GC, since we do not a GC phase on one node
to be prevented from cleaning up its RBT because a GC phase on a different node threw away some blocks
that were needed.
Line \ref{substitute} says what to do in this case.

### 2023-01-11 podc_paper/correctness-GC.tex
final edits

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $\var{v}.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in \var{v}'s tree at some time during the execution.
It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
each time a new block appears in \var{v}'s RBT, its \fld{index} is greater than any block
that has appeared in \var{v}'s RBT earlier.
Thus, $\var{v}.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.
We describe later what should happen if that block is not found by the search because it has been removed
by a GC phase.
\here{remember to come back to this point}

We now establish that definition (\ref{defsubblock}) of a block's subblocks
still makes sense by proving the analogue of Lemma \ref{lem::headProgress}.

@@

\begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
 If $h>0$ and a block with index $h$ has been inserted into $\var{v}.\fld{blocks}$ then 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{left}}$ and 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{right}}$.
\end{customlemma}
\begin{proof}
The block $B$ with index $h$ was installed into \var{v}'s RBT by the CAS at line \ref{casGC} 
that changed the tree from $T$ to $T'$.
Before this CAS, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
and then line \ref{invokeCreateBlockGC} created the block $B$ with $\fld{index}=h$.
Since $B'$ was already in $T$ before $B$ was created, the
\op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the 
\op{CreateBlock}$(v,b)$ that created $B$ started.
By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ 
stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.  
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

\Cref{lem::headProgressGC} implies that 
the nodes of an in-order traversal of any RBT have non-decreasing values of
$\eleft$ (and of $\eright$).
Thus, the searches for a block based on \eleft\ or \eright\ values 
at lines \ref{searchsuper1}, \ref{searchsuper2} and \ref{searchsuper3},
which are used to look for the superblock of a node or its predecessor, can be done in logarithmic time.

@@

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations of $\var{root}.\fld{blocks}[1..m]$.
Between the termination of $A$'s call to \op{Splitblock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates, the $A$ helps all pending dequeues at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks, so the claim is true.

@@

\here{Should add some discussion about the fact that we keep the newest finished block, in case some operation contained in a block with index i also needs to look at fields of the block with index i-1}

@@

Similarly the results of Section \ref{sec::linearizability} 
can be reproved in exactly the same way as for the original algorithm to establish
that the space-bounded algorithm is linearizable.

@@

We remark that the implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
Since each operation still only does two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.

>>>>

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $\var{v}.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in \var{v}'s tree at some time during the execution.
It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
each time a new block appears in \var{v}'s RBT, its \fld{index} is greater than any block
that has appeared in \var{v}'s RBT earlier.
Thus, $\var{v}.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.

We now establish that Definition (\ref{defsubblock}) of a block's subblocks
still makes sense by proving the analogue of Lemma \ref{lem::headProgress}.

@@

\begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
 If $h>0$ and a block with index $h$ has been inserted into $\var{v}.\fld{blocks}$ then 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{left}}$ and 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{right}}$.
\end{customlemma}
\begin{proof}
The block $B$ with index $h$ was installed into \var{v}'s RBT by the \op{CAS} at line \ref{casGC}.
Suppose that \op{CAS} changed the tree from $T$ to $T'$.
Before this \op{CAS}, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
and then line \ref{invokeCreateBlockGC} created the block $B$ with $\fld{index}=h$.
Since $B'$ was already in $T$ before $B$ was created, the
\op{CreateBlock}$(\var{v},b-1)$ that created $B'$ terminated before the 
\op{CreateBlock}$(\var{v},b)$ that created $B$ started.
By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(\var{v},b-1)$ 
stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
of \op{CreateBlock}$(\var{v},b)$ stores in $B.\eleft$.  
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

\Cref{lem::headProgressGC} implies that 
the nodes of an in-order traversal of any RBT have non-decreasing values of
$\eleft$ (and of $\eright$).
Thus, the searches for a block based on \eleft\ or \eright\ values 
at lines \ref{searchsuper3}, \ref{searchsuper1} and \ref{searchsuper2},
which are used to look for the superblock of a node or its predecessor, can be done in logarithmic time.

@@

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations of $\var{root}.\fld{blocks}[1..m]$.
Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates, the $A$ helps all pending dequeues at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks, so the claim is true.

@@

We remark that \Cref{lem::finished} guarantees that GC keeps one block that is finished and discards all blocks with smaller indices.
This is because the first unfinished block may still be traversed by an operation in the future,
and when examining that block the operation may need information from the preceding block.  For example,
when \op{FindResponse} is called on a block with index $b$, line \ref{FRnumenqGC} looks up the block with
index $b-1$.

@@

Similarly, the results of Section \ref{sec::linearizability} 
can be reproved in exactly the same way as for the original algorithm to establish
that the space-bounded algorithm is linearizable.

@@

The implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
Since each operation still only does two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.
